**Accuracy and precision of 16S rRNA gene sequence assignments to OTUs is more important than the stability of OTUs**

### 2015-10-01 Schloss_Cluster_PeerJ_2015.Rmd
Working on R&D

***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets. These algorithms utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database. This means that the best possible match is not necessarily returned once a highly similar hit is obtained. In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.

 This also explains the variation observed in the AGC and DGC algorithms, which rely upon the USEARCH algorithm.
[   Something, something, something.    ]

@@

***Problems with reference-based clustering.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs. This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. The default QIIME database, which was used in the He study and by others, contains full-length sequences that are at most 97% similar to each other. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1], 2)` sequences that mapped to a reference OTU per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`). The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized. This surprising result was likely due to the performance of the USEARCH heuristic. As with the AGC and DGC algorithms, if the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points. To test this further, we substituted VSEARCH for USEARCH in the closed-reference algorithm. When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs. That VSEARCH mapped more sequences than even the maximum number of sequences found across the USEARCH randomizations supports the VSEARCH developers' assertion that it has greater sensitivity than USEARCH.

@@

v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)
n_v4_overlap
n_v4_unique <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))

v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")

@@

***Roesch Canadian soil dataset***
After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length. Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan("data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan("data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences. Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set. The random sampling was repeated 30 times. The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets. For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters. Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.XX) as appropriate.

>>>>

***Evalution of an open-source alternative to USEARCH.*** The AGC and DGC algorithms appear to perform as well and possibly than the hierarchical clustering algorithms for some datasets. These algorithms utilize the USEARCH program (REF), which is not available as open source code and is only available for free to academic users as a 32-bit program. Access for non-academic users and those needing the 64-bit version is available commercially from the developer. An alternative to USEARCH is VSEARCH, which is being developed in parallel to USEARCH as a open-source alternative. One subtle difference between the two programs is that USEARCH employs a heuristic to determine when to stop searching a database because it think it has found the best hit or does not believe that the database contains a hit. This means that the best possible match is not necessarily returned once a highly similar hit is obtained. In contrast, VSEARCH does not utilize the heuristic, which the VSEARCH developers claim enhances the sensitivity relative to USEARCH. Using the two datasets, we determined whether the AGC and DGC algorithms as implemented by the two methods yielded sequencing assignments of similar quality.

This also explains the variation observed in the AGC and DGC algorithms, which rely upon the USEARCH algorithm.

@@

***Problems with reference-based clustering in general and as implemented in QIIME.*** The He study indicated that the closed-reference algorithm generated perfectly stable OTUs. This was unsurprising since, by definition, the algorithm represents a one-to-one mapping of reads to a reference and treats the input sequences independently. An important test that was not performed in the He study was to determine whether the clustering was sensitive to the order of the sequences in the database. The default QIIME database, which was used in the He study and by others, contains full-length sequences that are at most 97% similar to each other. We randomized the order of the reference sequences 30 times and used them to carry out the closed-reference algorithm with the full MiSeq dataset, which contained `r n_unique_miseq_seqs` unique sequences. Surprisingly, we observed that the number of OTUs generated was not the same in each of the randomizations. On average there were `r round(rand_ref_u["mean", 1], 2)` sequences that mapped to a reference OTU per randomization (range from `r round(rand_ref_u["min", 1], 2)` to `r round(rand_ref_u["max", 1], 2)`). The original ordering of the reference resulted in `r rand_ref_u["original", 1]` sequences being mapped, less than the minimum observed number of mapped sequences when the references were randomized. This surprising result was likely due to the performance of the USEARCH heuristic. As with the AGC and DGC algorithms, if the order of the reference sequences differed between runs, then the heuristic criteria would be triggered at different points. To test this further, we substituted VSEARCH for USEARCH in the closed-reference algorithm. When we used VSEARCH the original ordering of the reference sequences and all randomizations were able to map `r rand_ref_v["original", 1]` sequences to reference OTUs. That VSEARCH mapped more sequences than even the maximum number of sequences found across the USEARCH randomizations supports the VSEARCH developers' assertion that it has greater sensitivity than USEARCH.

@@

v19_summary <- read.table(file="data/gg_13_8/gg_13_8_97.v19.summary", header=T)

n_v4_unique <- length(scan("data/gg_13_8/gg_13_8_97.v4_ref.names", what="", sep="\n", quiet=T))

u_overlap <- read.table(file="data/rand_ref/overlap.uclosed.summary")
v_overlap <- read.table(file="data/rand_ref/overlap.vclosed.summary")

@@

***Roesch Canadian soil dataset***
After obtaining the 16S rRNA gene fragments from GenBank (accessions EF308591-EF361836), we followed the methods outlined by the He study by removing any sequence that contained an ambiguous base, was identified as being a chimera, and fell outside a defined sequence length. Although they reported observing a total of 50,542 sequences that were represented by 13,293 unique sequences, we obtained a total of `r length(scan("data/he/canada_soil.good.unique.pick.redundant.fasta", what="", sep="\n", quiet=T))/2` sequences that were represented by `r length(scan("data/he/canada_soil.good.unique.pick.fasta", what="", sep="\n", quiet=T))/2` unique sequences. Similar to the He study, we randomly sampled, without replacement, 20, 40, 60, and 80% of the sequences from the full data set. The random sampling was repeated 30 times. The order of the sequences in the full dataset was randomly permuted without replacement to generate an additional 30 datasets. For the hierarchical clustering algorithms and to generate a distance matrix, the pairwise distances between sequences were calculated in mothur using the pairwise.seqs command with the default Needleman-Wunsch alignment algorithm and parameters. Execution of the clustering algorithms was performed as described in the original He study using mothur (v.1.37) and QIIME (v.1.9.1) as appropriate.
